### 2025-09-18 paper/sec/introduction.tex
Progress on intro.

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making it hard for reviewers to check
them directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but currently
LLM interfaces do not support direct interrogation of visual outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, approaches the transparency problem from this direction, through software infrastructure which
transparently connections charts and figures to their data sources. \rpnote{Contrast with LLM-based approach
-- trusted infrastructure.} This enables visual outputs to provide interactive \emph{provenance queries}, user
interactions (e.g. mousing over a visual element) that allow readers to explore how visual features relate to
data. The value proposition of these approaches is that the interactions are provided automatically via
trusted infrastructure, namely an underlying programming language (say query language or general-purpose
language) which tracks dependencies as data flows through a computation. However, these approaches are limited
to outputs computed from data, which usually precludes natural language, despite the fact that the main claims
of a paper are usually made using natural language, with charts and figures only playing a supporting role.

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making it hard for reviewers to check
them directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, takes a more infrastructural approach to the transparency problem, by linking visualisations to their
data sources directly. This allows visual outputs to support in situ \emph{provenance queries}, user
interactions (e.g. mousing over a visual element) that reveal how visual features relate to data. The value
proposition of these approaches is that the interactions are provided automatically via trusted
infrastructure, namely a query language or general-purpose programming language which tracks dependencies as
data flows through a computation. However, these approaches to transparency only support direct interrogation
of outputs computed from data, such as visualisations. The accompanying natural language, which often plays a
central role in making the claims of the paper, remains hard to interpret.
